cubble: An R Package for Structuring Spatio-temporal Data

### 2022-04-26 arxiv/cubble-arxiv.tex
Some rewriting and comments added

\title{cubble An R Package for Structuring Spatio-temporal Data}

@@

\begin{abstract}
Spatio-temporal data refer to measurements taken across space and time. In practice, spatio-temporal data can be decomposed into a spatial and temporal component: at one time, we would select a spatial location and inspect the temporal trend; at other time, we might select one or multiple time value(s) and explore the spatial distribution. Ideally, we could make multiple maps and multiple time series to explore these together. However, doing all of these actions is complicated when data arrive fragmented in multiple objects. To make it easy to do all these tasks, ideally, spatial and temporal variables are in a single data object that we can slice and dice in different ways to conduct different visualisations. This work suggests a new data structure, \code{cubble}, that uses a nested form and a long form to organise spatial and temporal variables in a single data object. The new data structure ensures that data in the two forms are synchronised while being flexible to explore the two aspects of spatio-temporal data. It also provides the capability to handle data with hierarchical structure, match data from multiple sources, construct interactive graphics, and perform spatio-temporal transformation. The paper will demonstrate \pkg{cubble} with examples using Australian climate weather stations, river level data, and climate reanalysis (ERA5) data.
\end{abstract}

\keywords{
    spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis
  }

@@

The SpatioTemproal CRAN task view (\protect\hyperlink{ref-ctvspatiotemporal}{Edzer Pebesma 2022}) has a section \emph{Representing data} that lists existing spatio-temporal data representation in \proglang{R}. Among them, E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}) proposes four spatio-temporal layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} (\protect\hyperlink{ref-spacetime}{E. Pebesma 2012}) package with the underlying spatial and temporal class: \pkg{sp} (\protect\hyperlink{ref-sp}{E. Pebesma and Bivand 2005}) and \pkg{xts} (\protect\hyperlink{ref-xts}{Ryan and Ulrich 2020}). The \pkg{stars} (\protect\hyperlink{ref-stars}{E. Pebesma 2021}) package has been a recent implementation to handle both raster and vector data using spatio-temporal arrays. It interfaces with \pkg{sf} (\protect\hyperlink{ref-sf}{E. J. Pebesma 2018}), which adopts the simple features representation of spatial data, and \pkg{tidyverse} (\protect\hyperlink{ref-tidyverse}{Wickham et al. 2019}), a suite of tools to wrangle and visualise data in \proglang{R}.

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the representations above, and fitting the raw data into a data object can sometimes be not easy. More often, spatio-temporal data are collected in separate 2D tables and analysts need to ensemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) map data records the shape of a collection of area of interest; 2) geo-scientific data records the longitude and latitude coordinates of locations in the area along with other meta data relates to the location, and; 3) temporal variables indexed by both location and time.

The other reason is how tidy data (\protect\hyperlink{ref-tidydata}{Wickham 2014}) should be applied to spatio-temporal data. The tidy data concept advocates the data to be structured into 1) one row per observation, 2) one column per variable, and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e.~\pkg{dplyr} (\protect\hyperlink{ref-dplyr}{Wickham et al. 2022}) and \pkg{ggplot2} (\protect\hyperlink{ref-ggplot2}{Wickham 2016}) for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over daily or sub-daily data over years, which are extensively collected and handled in time series analysis. This poses the question of how to arrange spatial and temporal variables that makes analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble}, for working with a new spatio-temporal data representation, which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. Among the four spacetime layouts in E. Pebesma (\protect\hyperlink{ref-spacetime}{2012}), \pkg{cubble} can be applied to full grid layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section \ref{cube} discusses the spatio-temporal cube to conceptualise spatio-temporal data. Section \ref{cubble} presents the main design and functionality of \pkg{cubble}. Section \ref{others} explains how cubble deals with more advanced considerations, including data with the hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \ref{examples} uses Australia weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example of how \pkg{cubble} handles NetCDF data is also provided. Section \ref{conclude} concludes the paper.

@@

Spatio-temporal data can be conceptualised using a cubical data model with three axes. This abstraction can be useful for generalising operations and visualisation: Lu, Appel, and Pebesma (\protect\hyperlink{ref-lu_multidimensional_2018}{2018}) shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; Bach et al. (\protect\hyperlink{ref-bach_review_2014}{2014}) reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article ``does not need to involve spatial data'', but refers to ``an abstract 2D substrate that is used to visualize data at a specific time''. Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisation still applies to our spatio-temporal data manipulation and visualisation.

While a cubical data model is neat to view the spatio-temporal data conceptually, a 3D data representation, or an array, may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easy done with array objects. 2) The \texttt{mutate()} syntax for creating new variables is no longer valid since the operation can no longer define how a created 1D vector should be added to the array.

The most common data cube is defined using the three axes: time, latitude, and longitude, which can be seen as taking snapshots of the space and stacking on time. Our spatio-temporal cube is defined slightly differently with three axes: time, site, and variable as Figure \ref{fig:cube-def}. The time axis is the same across the two formations, the site encompasses latitude and longitude into a single column identifier, while variables are stacked on this defined space-time canvas. This notion is adopted to avoid using hypercube when describing multivariate spatio-temporal data.

@@

{\centering \includegraphics[width=0.5\linewidth]{/Users/sherryzhang/Documents/research/paper-cubble/figures/cube-def}

@@

\caption{An ilustration of the conceputalised spatio-temporal cube with three axies: time, site, and variable.}\label{fig:cube-def}
\end{figure}

@@

This section will first introduce the core functions in the cubble class: \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()} in subsection \ref{create} to \ref{unfold}. The next subsection addresses the necessity of creating a new class. Subsection \ref{compact} will then show the compatibility of cubble with existing packages in spatial and temporal analysis.

Each core function is introduced with a short example accompanied and the data \texttt{climate\_flat} is used throughout this section. \texttt{climate\_flat} contains five weather stations in Australia with spatial information of each station: station id, latitude, longitude, elevation, station name, and World Meteorology Organisation ID and also daily temporal information of date, maximum and minimum temperature and precipitation for 2020. Below prints out the first five rows of \texttt{climate\_flat}:

@@

Spatio-temporal data can come in various formats and shapes and a cubble can be created from various raw data. This includes tibble and its variates multiple tables (an example provided in section \ref{covid}), and netcdf (detailed in section \ref{netcdf}). The function \code{as_cubble()} is used to create a cubble with three additional components: \code{key} as the spatial identifier; \code{index} as the temporal identifier; and a vector of \code{coords} in the order of longitude and latitude. Argument \code{key} and \code{index} follow the wording in \pkg{tsibble} to describe the temporal order and multiple series while \code{coords} is a space specific argument to locate each site in two columns. The code below creates a cubble out of \code{climate_flat} with \code{id} as the key, \code{date} as the index, and \code{c(long, lat)} as the coordinates:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_nested }\OtherTok{\textless{}{-}}\NormalTok{ climate\_flat }\SpecialCharTok{|\textgreater{}}
  \FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date, }\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}
\NormalTok{cubble\_nested}
\end{Highlighting}
\end{Shaded}

@@

The cubble header provides some information about this data. \code{id} is the variable name to identify each location and there are five unique locations. The bounding box is \code{[115.97, -32.94, 133.55, -12.42]}. In the third row is the the name and type of variable nested in the \code{ts} column and here it includes \code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

The created cubble is built from a \code{rowwise_df} class where each row forms a group. All the temporal variables are nested in a list column, hence it is also called the nested cubble. The rowwise structure makes it simpler to calculate on the list using the \code{mutate()} syntax, which is simpler than the \{\code{map()} when working with list column. For example, calculating the number of raining days can be done by:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_nested }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{rain\_day =} \FunctionTok{sum}\NormalTok{(ts}\SpecialCharTok{$}\NormalTok{prcp }\SpecialCharTok{!=} \DecValTok{0}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{change-focus-to-the-time-variable-face}{%
\subsection{Change focus to the time-variable face}\label{change-focus-to-the-time-variable-face}}

The nested form can be used for those operations whose output is only indexed by the spatial identifier (\code{key}), while for those outputs cross-identified by the spatial and temporal identifier (\code{key} and \code{index}), the nested form becomes inadequate. \pkg{cubble} also provides a long form, which expands the \code{ts} column and temporarily ``hide'' the spatial variables. The function \code{face_temporal()} is used to switch a nested cubble into a long one and the first row in Figure \ref{fig:face} illustrates this operation where the focus of the cube now changes from the site-variable face to the time-variable face. This code switches the cubble just created into its long form:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_long }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_nested }\SpecialCharTok{|\textgreater{}} \FunctionTok{face\_temporal}\NormalTok{()}
\NormalTok{cubble\_long}
\end{Highlighting}
\end{Shaded}

@@

The first line in the header now shows the cubble is in its long form and the third line has been changed to show the name and type of spatial variables: \code{lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]}. Unlike the nested form, the long cubble is built form class \code{groupped_df} where all the observations from the same sites form a group.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.008}

@@

\caption{An illustration of function \code{face\_temporal} and \code{face\_spatial} in cubble. In the first row, \code{face\_temporal} switches a cubble from the nested form into the long form and the focus has switched from the spatial aspect (the side face) to the temporal aspect (the front face). In the second row, \code{face\_spatial} switches a cubble back to the nested form from te long form and shift focus back to the spatial aspect.}\label{fig:face}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_back }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_long }\SpecialCharTok{|\textgreater{}} \FunctionTok{face\_spatial}\NormalTok{()}
\NormalTok{cubble\_back}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{unfold}{%
\subsection{Flatten spatial variables into the long cubble}\label{unfold}}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_unfold }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_long }\SpecialCharTok{|\textgreater{}} \FunctionTok{unfold}\NormalTok{(long, lat)}
\NormalTok{cubble\_unfold}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_unfold }\SpecialCharTok{|\textgreater{}} \FunctionTok{face\_spatial}\NormalTok{() }\SpecialCharTok{|\textgreater{}} \FunctionTok{face\_temporal}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

@@

\subsection[Why not just use]{Why not just use \code{dplyr::nest\_by()}}

Some readers may raise the question why a new data structure rather than using \code{dplyr::nest_by()} on the combined data to directly create a list-column. While \code{nest_by()} and its pair \code{unnest()} can mimic the nested form in cubble, cubble can be seen as an attempt to arrange variables from different observational units into a single object, specifically for spatio-temporal data. While \texttt{nest\_by()} can recreate the nested form, it can't create a clean slate for temporal variables as the long form. For temporal variables, carrying the duplicated spatial variables in the combined form is not convenient to work with. This is because these spatial variables will lose after \code{summarise()} and one will need to subset the spatial variables again from the original data and make a left join, which tends to create unnecessary intermediate products in the workflow.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# operation on the long form w/o cubble}
\CommentTok{\# cubble workflow}
\NormalTok{dt }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{as\_cubble}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{face\_temporal}\NormalTok{() }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{group\_by}\NormalTok{() }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{summarise}\NormalTok{() }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{unfold}\NormalTok{(...)}

\CommentTok{\# tidyverse workflow}
\NormalTok{temporal }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{group\_by}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}} 
  \FunctionTok{summarise}\NormalTok{(...)}

\NormalTok{spatial }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{|\textgreater{}} \FunctionTok{select}\NormalTok{(....)}

\NormalTok{spatial }\SpecialCharTok{|\textgreater{}} \FunctionTok{left\_join}\NormalTok{(temporal)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# example with a tsibble created from climate\_flat}
\NormalTok{raw }\OtherTok{\textless{}{-}}\NormalTok{ climate\_flat }\SpecialCharTok{|\textgreater{}}\NormalTok{ tsibble}\SpecialCharTok{::}\FunctionTok{as\_tsibble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date)}
\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{  raw }\SpecialCharTok{|\textgreater{}}\NormalTok{ cubble}\SpecialCharTok{::}\FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}
\NormalTok{dt}
\end{Highlighting}
\end{Shaded}

@@

In the nested cubble created, each element in the list-column \code{ts} is of \code{tbl_ts} class and operations available to the tsibble class is still valid under cubble. For example, the code below calculates two features of the maximum temperature:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# add station{-}based features in the nested form.}
\NormalTok{dt }\SpecialCharTok{|\textgreater{}} \FunctionTok{mutate}\NormalTok{(fabletools}\SpecialCharTok{::}\FunctionTok{features}\NormalTok{(ts, tmax, }\FunctionTok{list}\NormalTok{(}\AttributeTok{tmax\_mean =}\NormalTok{ mean, }\AttributeTok{tmax\_var =}\NormalTok{ var)))}
\end{Highlighting}
\end{Shaded}

@@

As a spatial data object, \code{sf} creates a simple feature geometry list-column (\code{sfc}) in the data frame to provide spatial operations on various geometry types (\code{POINT}, \code{LINESTRING}, \code{POLYGON}, \code{MULTIPOLYGON}, etc). An \code{sf} object can be supplied as the spatial table, as part of the raw data, along with a temporal table, to create a cubble. Once formed in a cubble, methods for the \code{sfc} class can be applied in the nested form. An example of this is shown in \ref{covid}, which also handles the case when the sites in the two tables don't match exactly. A spatial data object with \code{s2} vector can also be the input of the spatial table.

>>>>

\title{cubble: An R Package for Structuring Spatio-temporal Data}

@@

\begin{abstract}
Spatio-temporal data refer to measurements taken across space and time. In practice, spatio-temporal data can be decomposed into a spatial and temporal component: one can select a spatial location and inspect the temporal trend; or, one might select one or multiple time value(s) and explore the spatial distribution. Ideally, multiple maps and multiple time series could be made to explore these together. However, doing all of these actions is complicated when data arrive fragmented in multiple objects. To make it easy to do all these tasks, ideally, spatial and temporal variables should be in a single data object that we can slice and dice in different ways to conduct different visualisations. This work suggests a new data structure, \code{cubble}, that uses a nested form and a long form to organise spatial and temporal variables in a single data object. The new data structure ensures that data in the two forms are synchronised while being flexible to explore the two aspects of spatio-temporal data. It also provides the capability to handle data with hierarchical structure, match data from multiple sources, construct interactive graphics, and perform spatio-temporal transformation. The paper will demonstrate the use of \pkg{cubble} with examples using Australian climate weather stations, river level data, and climate reanalysis (ERA5) data.
\end{abstract}

\keywords{
    spatial, temporal, spatio temporal, R, exploratory data analysis, environmental data, climate data
  }

@@

The SpatioTemproal CRAN task view (\protect\hyperlink{ref-ctvspatiotemporal}{Edzer Pebesma 2022}) has a section \emph{Representing data} that lists existing spatio-temporal data representations in \proglang{R}. Among them, \protect\hyperlink{ref-spacetime}{E. Pebesma} (\protect\hyperlink{ref-spacetime}{2012}) proposes four spatio-temporal layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} (\protect\hyperlink{ref-spacetime}{E. Pebesma 2012}) package with the underlying spatial and temporal classes \pkg{sp} (\protect\hyperlink{ref-sp}{E. Pebesma and Bivand 2005}) and \pkg{xts} (\protect\hyperlink{ref-xts}{Ryan and Ulrich 2020}). The \pkg{stars} (\protect\hyperlink{ref-stars}{E. Pebesma 2021}) package has been a recent implementation to handle both raster and vector data using spatio-temporal arrays. It interfaces with \pkg{sf} (\protect\hyperlink{ref-sf}{E. J. Pebesma 2018}), which adopts the simple features representation of spatial data, and \pkg{tidyverse} (\protect\hyperlink{ref-tidyverse}{Wickham et al. 2019}), a suite of tools to wrangle and visualise data in \proglang{R}.

Still, the data representation for spatio-temporal data can be further improved and there are two reasons for this. Firstly, the raw data sourced in the wild is less often presented in any one of the representations above, and fitting the raw data into a data object can sometimes be difficult. More often, spatio-temporal data are collected in separate 2D tables and analysts need to assemble them into a whole piece before exploring the data. Examples of components of spatio-temporal data can be 1) map data recording the shape of a collection of areas of interest; 2) geo-scientific data recording the longitude and latitude coordinates of locations in the area along with other metadata related to the location, and; 3) temporal variables indexed by both location and time.

The other reason is how tidy data concepts (\protect\hyperlink{ref-tidydata}{Wickham 2014}) should be applied to spatio-temporal data. Tidy data should be structured into 1) one row per observation, 2) one column per variable, and 3) one type of data per table. The long form data is preferred over wide data given the downstream software i.e.~\pkg{dplyr} (\protect\hyperlink{ref-dplyr}{Wickham et al. 2022}) and \pkg{ggplot2} (\protect\hyperlink{ref-ggplot2}{Wickham 2016}) for data wrangling and visualisation. However, the long form can be inefficient to store feature geometries, especially for large multipolygons for daily or sub-daily data over years, which are extensively collected and handled in time series analysis. This poses the question of how to arrange spatial and temporal variables in a way that would make analysing spatio-temporal data easier.

This paper presents a new \proglang{R} package, \pkg{cubble} which addresses the two issues mentioned above. In the package, a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while being kept synchronised. Among the four spacetime layouts in \protect\hyperlink{ref-spacetime}{E. Pebesma} (\protect\hyperlink{ref-spacetime}{2012}), \pkg{cubble} can be applied to full grid layout, sparse grid layout, or irregular layout, but not trajectory, which is outside the scope of this work. The software is available from the Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section \ref{cube} introduces the proposed cube structure as a way to conceptualise multivariate spatio-temporal data. Section \ref{cubble} presents the main design and functionality of \pkg{cubble}. Section \ref{others} explains how cubble deals with more advanced considerations, including data with the hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \ref{examples} uses Australian weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example of how \pkg{cubble} handles NetCDF data is also provided. Section \ref{conclude} concludes the paper.

@@

\textcolor{blue}{Ursula: I have rearranged this section somewhat, please check!}

Spatio-temporal data can be conceptualised using a cubical data model with three axes, typically time, latitude and longitude. This abstraction can be useful for generalising operations and visualisation: \protect\hyperlink{ref-lu_multidimensional_2018}{Lu, Appel, and Pebesma} (\protect\hyperlink{ref-lu_multidimensional_2018}{2018}) shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; \protect\hyperlink{ref-bach_review_2014}{Bach et al.} (\protect\hyperlink{ref-bach_review_2014}{2014}) reviews the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article ``does not need to involve spatial data,'' but refers to ``an abstract 2D substrate that is used to visualize data at a specific time.'' Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualisation still applies to our spatio-temporal data manipulation and visualisation.

The most common data cube using the three axes (time, latitude, longitude) could be considered as taking snapshots of the space and stacking on time. Here we will define the spatio-temporal cube slightly differently with the three axes (time, site, variable), as illustrated in Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hypercubes when describing multivariate spatio-temporal data.

While a cubical data model is neat to view the spatio-temporal data conceptually, a 3D data representation, or an array, may not be convenient for data wrangling. There are two reasons for this: 1) While arrays are efficient for the computation on numerical values, spatio-temporal data contains more than just numerical variables. For example, wrangling on character strings and specific datetime classes is common for spatio-temporal data but may not be easily done with array objects. 2) The \texttt{mutate()} syntax for creating new variables is no longer valid since the operation can no longer define how a created 1D vector should be added to the array.
\textcolor{blue}{Ursula: this seems too specific, you should be able to make the same argument without referencing the mutate function?}

@@

{\centering \includegraphics[width=0.5\linewidth]{/Users/ursulalaa/papers/paper-cubble/figures/cube-def}

@@

\caption{An illustration of the conceptualised spatio-temporal cube with three axes: time, site, and variable.}\label{fig:cube-def}
\end{figure}

@@

This section will first introduce the core functions in the cubble class: \code{as_cubble()}, \code{face_spatial()}, \code{face_temporal()}, and \code{unfold()} in subsection \ref{create} - \ref{unfold}. The next subsection addresses the necessity of creating a new class. Subsection \ref{compact} will then show the compatibility of cubble with existing packages in spatial and temporal analysis.

Each core function is introduced accompanied with a short example and the data \texttt{climate\_flat} is used throughout this section. \texttt{climate\_flat} contains five weather stations in Australia with spatial information of each station: station id, latitude, longitude, elevation, station name, World Meteorology Organisation ID and also daily temporal information of date, maximum and minimum temperature and precipitation for 2020. The first five rows of \texttt{climate\_flat} are shown below:

@@

Spatio-temporal data can come in various formats and shapes and a cubble can be created from various types of raw data. This includes tibble and its variates in multiple tables (an example is provided in section \ref{covid}), and netCDF (detailed in section \ref{netcdf}). The function \code{as_cubble()} is used to create a cubble with three additional inputs: \code{key} as the spatial identifier; \code{index} as the temporal identifier; and a vector of \code{coords} in the order (longitude, latitude). The arguments \code{key} and \code{index} follow the wording in \pkg{tsibble} to describe the temporal order and multiple series while \code{coords} specifies the spatial location of each site. The code below creates a cubble out of \code{climate_flat} with \code{id} as the key, \code{date} as the index, and \code{c(long, lat)} as the coordinates:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_nested }\OtherTok{\textless{}{-}}\NormalTok{ climate\_flat }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date, }\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}
\NormalTok{cubble\_nested}
\end{Highlighting}
\end{Shaded}

@@

The cubble header provides some information about this data. \code{id} is the variable name to identify each location and there are five unique locations. The bounding box is \code{[115.97, -32.94, 133.55, -12.42]}. The third row shows the name and type of all variables nested in the \code{ts} column and here it includes \code{date [date], prcp [dbl], tmax [dbl], tmin [dbl]}.

The resulting cubble is built from a \code{rowwise_df} class where each row forms a group. All the temporal variables are nested in a list column, hence it is also called the nested cubble. The rowwise structure makes it simpler to calculate on the list using the \code{mutate()} syntax, which is simpler than the \code{map()} when working with list column. For example, calculating the number of raining days can be done by:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_nested }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{rain\_day =} \FunctionTok{sum}\NormalTok{(ts}\SpecialCharTok{$}\NormalTok{prcp }\SpecialCharTok{!=} \DecValTok{0}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{change-focus-by-facing-the-time-variables}{%
\subsection{Change focus by facing the time-variables}\label{change-focus-by-facing-the-time-variables}}

The nested form can be used for those operations where the output is only indexed by the spatial identifier (\code{key}), but becomes inadequate when outputs need both a spatial and a temporal identifier (\code{key} and \code{index}). \pkg{cubble} also provides a long form, which expands the \code{ts} column and temporarily ``hides'' the spatial variables. The function \code{face_temporal()} is used to switch a nested cubble into a long one and the first row in Figure \ref{fig:face} illustrates this operation where the focus of the cube now changes from the site-variable face to the time-variable face. This code switches the cubble just created into its long form:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_long }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_nested }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_temporal}\NormalTok{()}
\NormalTok{cubble\_long}
\end{Highlighting}
\end{Shaded}

@@

The first line in the header now shows the cubble is in its long form and the third line has been changed to show the name and type of spatial variables: \code{lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]}. Unlike the nested form, the long cubble is built from class \code{groupped_df} where all the observations from the same sites form a group.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/ursulalaa/papers/paper-cubble/figures/diagram-keynotes/diagram-keynotes.008}

@@

\caption{An illustration of function \code{face\_temporal} and \code{face\_spatial} in cubble. In the first row, \code{face\_temporal} switches a cubble from the nested form into the long form and the focus has switched from the spatial aspect (the side face) to the temporal aspect (the front face). In the second row, \code{face\_spatial} switches a cubble back to the nested form from the long form and shifts focus back to the spatial aspect.}\label{fig:face}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_back }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_long }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_spatial}\NormalTok{()}
\NormalTok{cubble\_back}
\end{Highlighting}
\end{Shaded}

@@

\hypertarget{unfold}{%
\subsection{Unfold spatial variables into the long cubble}\label{unfold}}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_unfold }\OtherTok{\textless{}{-}}\NormalTok{ cubble\_long }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{unfold}\NormalTok{(long, lat)}
\NormalTok{cubble\_unfold}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cubble\_unfold }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_spatial}\NormalTok{() }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{face\_temporal}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

@@

\subsection[Why not just use]{Why not just use \code{dplyr::nest\_by()}?}

\textcolor{blue}{Ursula: I feel like this doesn't get the message accross, the main point seems to be that with cubble we can switch and never have to add information back in "by-hand" (with some join function), maybe make the point without refering to specific functions first and then explain the example?}

Some readers may raise the question why a new data structure rather than using \code{dplyr::nest_by()} on the combined data to directly create a list-column. While \code{nest_by()} and its inverse \code{unnest()} can mimic the nested form in cubble, cubble can be seen as an attempt to arrange variables from different observational units into a single object, specifically for spatio-temporal data. While \texttt{nest\_by()} can recreate the nested form, it cannot create a clean slate for temporal variables as used in the long form. For temporal variables, carrying the duplicated spatial variables in the combined form is not convenient to work with. This is because these spatial variables will be lost after \code{summarise()} and one will need to subset the spatial variables again from the original data and make a left join, which tends to create unnecessary intermediate products in the workflow.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# operation on the long form w/o cubble}
\CommentTok{\# cubble workflow}
\NormalTok{dt }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{as\_cubble}\NormalTok{(...) }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{face\_temporal}\NormalTok{() }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{group\_by}\NormalTok{() }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{summarise}\NormalTok{() }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{unfold}\NormalTok{(...)}

\CommentTok{\# tidyverse workflow}
\NormalTok{temporal }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{group\_by}\NormalTok{(...) }\SpecialCharTok{|}\ErrorTok{\textgreater{}}
  \FunctionTok{summarise}\NormalTok{(...)}

\NormalTok{spatial }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{select}\NormalTok{(....)}

\NormalTok{spatial }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{left\_join}\NormalTok{(temporal)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# example with a tsibble created from climate\_flat}
\NormalTok{raw }\OtherTok{\textless{}{-}}\NormalTok{ climate\_flat }\SpecialCharTok{|}\ErrorTok{\textgreater{}}\NormalTok{ tsibble}\SpecialCharTok{::}\FunctionTok{as\_tsibble}\NormalTok{(}\AttributeTok{key =}\NormalTok{ id, }\AttributeTok{index =}\NormalTok{ date)}
\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{  raw }\SpecialCharTok{|}\ErrorTok{\textgreater{}}\NormalTok{ cubble}\SpecialCharTok{::}\FunctionTok{as\_cubble}\NormalTok{(}\AttributeTok{coords =} \FunctionTok{c}\NormalTok{(long, lat))}
\NormalTok{dt}
\end{Highlighting}
\end{Shaded}

@@

In the nested cubble created, each element in the list-column \code{ts} is of \code{tbl_ts} class and operations available to the tsibble class are still valid within the cubble. For example, the code below calculates two features of the maximum temperature:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# add station{-}based features in the nested form.}
\NormalTok{dt }\SpecialCharTok{|}\ErrorTok{\textgreater{}} \FunctionTok{mutate}\NormalTok{(fabletools}\SpecialCharTok{::}\FunctionTok{features}\NormalTok{(ts, tmax, }\FunctionTok{list}\NormalTok{(}\AttributeTok{tmax\_mean =}\NormalTok{ mean, }\AttributeTok{tmax\_var =}\NormalTok{ var)))}
\end{Highlighting}
\end{Shaded}

@@

As a spatial data object, \code{sf} creates a simple feature geometry list-column (\code{sfc}) in the data frame to provide spatial operations on various geometry types (\code{POINT}, \code{LINESTRING}, \code{POLYGON}, \code{MULTIPOLYGON}, etc). When creating a cubble an \code{sf} object can be supplied as the spatial table. Once included in a cubble, methods for the \code{sfc} class can be applied in the nested form. An example of this is shown in \ref{covid}, which also handles the case when the site identifiers in the two tables do not match exactly. A spatial data object with \code{s2} vector can also be the input for the spatial table.
